---
title: "Citywide Crime Rate Data Set Creation, 2017-2019"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
params:
  months: 9
---

## Introduction
This notebook creates a table documenting changes in citywide crime rate by year.

## Dependencies
This notebook requires the following `R` packages:

```{r load-packages, message=FALSE}
# tidyverse packages
library(dplyr)
library(readr)
library(stringr)

# spatial packages
library(tidycensus)

# other packages
library(compstatr)
library(here)
library(testthat)
```

## Load Data
The data for this table will be drawn from the three yearly `.csv` files created by `01_CreateData_Crime.Rmd`. We'll load those first:

```{r load-data}
part1_2017 <- read_csv(here("data", "crimes", "part1_2017.csv"))
part1_2018 <- read_csv(here("data", "crimes", "part1_2018.csv"))
part1_2019 <- read_csv(here("data", "crimes", "part1_2019.csv"))
```

## Subset Data
Next, we'll subset these data into violent crime and homicide objects:

```{r subset-data-by-crime}
# 2017
homicide_2017 <- cs_filter_crime(part1_2017, var = crime, crime = "Homicide")
violent_2017 <- cs_filter_crime(part1_2017, var = crime, crime = "Violent")

# 2018
homicide_2018 <- cs_filter_crime(part1_2018, var = crime, crime = "Homicide")
violent_2018 <- cs_filter_crime(part1_2018, var = crime, crime = "Violent")

# 2019
homicide_2019 <- cs_filter_crime(part1_2019, var = crime, crime = "Homicide")
violent_2019 <- cs_filter_crime(part1_2019, var = crime, crime = "Violent")
```

To make direct comparisons between the 2019 data and the earlier years, we need to create partial year objects for 2017 and 2018:

```{r subset-data-by-date}
# define last dates
last17 <- paste0("2017-", 
                 str_pad(as.character(params$months+1), width = 2, side = "left", pad = "0"), 
                 "-01")
last18 <- paste0("2018-", 
                 str_pad(as.character(params$months+1), width = 2, side = "left", pad = "0"), 
                 "-01")

# 2017
homicide_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> homicide_2017p
  
violent_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> violent_2017p

part1_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> part1_2017p

# 2018
homicide_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> homicide_2018p
  
violent_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> violent_2018p

part1_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> part1_2018p
```

## Create City-wide Rates
First, we want to create a tibble with just summary data:

```{r create-city-counts}
city <- tibble(
  year = c(2017, 2017, 2017, 2017, 2017, 2017,
           2018, 2018, 2018, 2018, 2018, 2018,
           2019, 2019, 2019),
  range = c("full", "full", "full", "partial", "partial", "partial",
            "full", "full", "full", "partial", "partial", "partial",
            "partial", "partial", "partial"),
  type = c("homicide", "part 1", "violent", "homicide", "part 1", "violent",
           "homicide", "part 1", "violent", "homicide", "part 1", "violent",
           "homicide", "part 1", "violent"),
  count = c(nrow(homicide_2017), nrow(part1_2017), nrow(violent_2017),
            nrow(homicide_2017p), nrow(part1_2017p), nrow(violent_2017p),
            nrow(homicide_2018), nrow(part1_2018), nrow(violent_2018),
            nrow(homicide_2018p), nrow(part1_2018p), nrow(violent_2018p),
            nrow(homicide_2019), nrow(part1_2019), nrow(violent_2019))
)
```

Next, we'll download the city's total population using the 2013-2017 ACS 5-year estimate:

```{r census-data-city, results="hide"}
# download data
county <- get_acs(year = 2017, geography = "county", variables = "B01003_001", 
                state = 29, county = 510, geometry = FALSE)

# store population value
county_pop <- county$estimate[1]
```

Finally, we'll add the population and then calculate rates:

```{r calculate-rates}
city %>%
  mutate(total_pop = county_pop) %>%
  mutate(rate = count/total_pop*1000) %>%
  select(year, range, type, count, rate, total_pop) -> city
```

## Write Data
Now we can write our city-wide estimates:

```{r write-data}
write_csv(city, path = here("data", "city", "city_wide.csv"))
```
