---
title: "Regional Crime Data, 2017-2019"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
params:
  months: 9
---

## Introduction
This notebook creates a set of `.csv` files containing estimates at the "regional" level - several experimental groupings of City of St. Louis neighborhoods to make direct comparisons to Hayden's Rectangle. 

## Dependencies
This notebook requires the following `R` packages:

```{r load-packages, message=FALSE}
# tidyverse packages
library(dplyr)
library(readr)
library(stringr)

# spatial packages
library(areal)
library(sf)
library(tidycensus)

# other packages
library(compstatr)
library(here)
library(testthat)
```

## Load Data
We have a shapefile containing these regions in the `data/raw` subdirectory:

```{r load-regions-data}
regions <- st_read(here("data", "raw", "haydenRegions"), stringsAsFactors = FALSE) %>%
  st_transform(crs = 26915) %>%
  select(-Shape_Leng, -Shape_Area)
```

The data for these table will be drawn from the three yearly `.csv` files created by `01_CreateData_Crime.Rmd`. We'll load those first:

```{r load-data}
part1_2017 <- read_csv(here("data", "crimes", "part1_2017.csv"))
part1_2018 <- read_csv(here("data", "crimes", "part1_2018.csv"))
part1_2019 <- read_csv(here("data", "crimes", "part1_2019.csv"))
```

## Prepare Regional Data
Our first task is to calculate the total estimated population for each region.

### Census Data
To calculate rates, we'll use total population estimates from the 2013-2017 5 year American Community Survey estimates. These data are downloaded along with their geometry, and subset down to just the GEOID value and the estimated population:

```{r census-data-tracts, results="hide"}
pop <- get_acs(year = 2017, geography = "tract", variables = "B01003_001", 
                state = 29, county = 510, geometry = TRUE) %>%
  st_transform(crs = 26915) %>%
  select(GEOID, estimate) %>%
  rename(total_pop = estimate)
```

### Estimate Neighborhood Region Population
With both the Census data and the regional geometry, we can use a technique called areal weighted interpolation to produce estimated populations for each region based on the degree of overlap between regional boundaries and census tract boundaries:

```{r interpolate}
# interpolate values
regions <- aw_interpolate(regions, tid = ID, source = pop, sid = GEOID, 
                        weight = "sum", output = "tibble", extensive = "total_pop")

# unit test estimates
expect_equal(aw_verify(source = pop, sourceValue = total_pop, 
                       result = regions, resultValue = total_pop), TRUE)
```

### Clean-up Enviornment
Next, we'll remove the `pop` object from our enviornment:

```{r clean-up-1}
rm(pop)
```

## Subset Data
Next, we'll subset these data into violent crime and homicide objects:

```{r subset-data-by-crime}
# 2017
homicide_2017 <- cs_filter_crime(part1_2017, var = crime, crime = "Homicide")
violent_2017 <- cs_filter_crime(part1_2017, var = crime, crime = "Violent")

# 2018
homicide_2018 <- cs_filter_crime(part1_2018, var = crime, crime = "Homicide")
violent_2018 <- cs_filter_crime(part1_2018, var = crime, crime = "Violent")

# 2019
homicide_2019 <- cs_filter_crime(part1_2019, var = crime, crime = "Homicide")
violent_2019 <- cs_filter_crime(part1_2019, var = crime, crime = "Violent")
```

To make direct comparisons between the 2019 data and the earlier years, we need to create partial year objects for 2017 and 2018:

```{r subset-data-by-date}
# define last dates
last17 <- paste0("2017-", 
                 str_pad(as.character(params$months+1), width = 2, side = "left", pad = "0"), 
                 "-01")
last18 <- paste0("2018-", 
                 str_pad(as.character(params$months+1), width = 2, side = "left", pad = "0"), 
                 "-01")

# 2017
homicide_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> homicide_2017p
  
violent_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> violent_2017p

part1_2017 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last17) %>%
  select(-dateTime, -date, -time) -> part1_2017p

# 2018
homicide_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> homicide_2018p
  
violent_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> violent_2018p

part1_2018 %>%
  cs_parse_date(var = date_occur, dateVar = date, timeVar = time) %>%
  filter(date < last18) %>%
  select(-dateTime, -date, -time) -> part1_2018p
```

## Summarize Data
With our objects created, we can group our data by region and them summarize them.

### 2017
First, homicides:

```{r homicides-2017}
# full
homicide_2017 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "full",
    type = "homicide"
  ) %>%
  select(year, range, type, region, count) -> homicide_2017

# partial
homicide_2017p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "partial",
    type = "homicide"
  ) %>%
  select(year, range, type, region, count) -> homicide_2017p
```

Next, violent crimes:

```{r violent-2017}
# full
violent_2017 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "full",
    type = "violent"
  ) %>%
  select(year, range, type, region, count) -> violent_2017

# partial
violent_2017p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "partial",
    type = "violent"
  ) %>%
  select(year, range, type, region, count) -> violent_2017p
```

Finally, Part 1 crimes:

```{r part1-2017}
# full
part1_2017 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "full",
    type = "part 1"
  ) %>%
  select(year, range, type, region, count) -> part1_2017

# partial
part1_2017p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2017,
    range = "partial",
    type = "part 1"
  ) %>%
  select(year, range, type, region, count) -> part1_2017p
```

With these created, we can combine them into a single 2017 object:

```{r bind-2017}
# bind
bind_2017 <- bind_rows(homicide_2017, part1_2017, violent_2017,
                       homicide_2017p, part1_2017p, violent_2017p)

# clean-up
rm(homicide_2017, part1_2017, violent_2017, homicide_2017p, part1_2017p, violent_2017p)
```

### 2018
First, homicides:

```{r homicides-2018}
# full
homicide_2018 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "full",
    type = "homicide"
  ) %>%
  select(year, range, type, region, count) -> homicide_2018

# partial
homicide_2018p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "partial",
    type = "homicide"
  ) %>%
  select(year, range, type, region, count) -> homicide_2018p
```

Next, violent crimes:

```{r violent-2018}
# full
violent_2018 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "full",
    type = "violent"
  ) %>%
  select(year, range, type, region, count) -> violent_2018

# partial
violent_2018p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "partial",
    type = "violent"
  ) %>%
  select(year, range, type, region, count) -> violent_2018p
```

Finally, Part 1 crimes:

```{r part1-2018}
# full
part1_2018 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "full",
    type = "part 1"
  ) %>%
  select(year, range, type, region, count) -> part1_2018

# partial
part1_2018p %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2018,
    range = "partial",
    type = "part 1"
  ) %>%
  select(year, range, type, region, count) -> part1_2018p
```

With these created, we can combine them into a single 2018 object:

```{r bind-2018}
# bind
bind_2018 <- bind_rows(homicide_2018, part1_2018, violent_2018,
                       homicide_2018p, part1_2018p, violent_2018p)

# clean-up
rm(homicide_2018, part1_2018, violent_2018, homicide_2018p, part1_2018p, violent_2018p)
```

### 2019
First, homicides:

```{r homicides-2019}
# partial
homicide_2019 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2019,
    range = "partial",
    type = "homicide"
  ) %>%
  select(year, range, type, region, count) -> homicide_2019
```

Next, violent crimes:

```{r violent-2019}
# partial
violent_2019 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2019,
    range = "partial",
    type = "violent"
  ) %>%
  select(year, range, type, region, count) -> violent_2019
```

Finally, Part 1 crimes:

```{r part1-2019}
# partial
part1_2019 %>%
  group_by(region) %>%
  summarize(count = n()) %>%
  mutate(
    year = 2019,
    range = "partial",
    type = "part 1"
  ) %>%
  select(year, range, type, region, count) -> part1_2019
```

With these created, we can combine them into a single 2019 object:

```{r bind-2019}
# bind
bind_2019 <- bind_rows(homicide_2019, part1_2019, violent_2019)

# clean-up
rm(homicide_2019, part1_2019, violent_2019)
```

## Combine and Calculate Rates
We can collapse our three objects further:

```{r collapse-again}
# bind
regional_crime <- bind_rows(bind_2017, bind_2018, bind_2019)

# clean-up
rm(bind_2017, bind_2018, bind_2019)
```

Then we can calculate rates:

```{r calculate-rates}
# calculate
regions %>%
  select(-ID) %>%
  rename(region = NAME) %>%
  left_join(regional_crime, ., by = "region") %>%
  mutate(rate = count/total_pop*1000) %>%
  select(year, range, type, region, count, rate, total_pop) -> regions

# clean-up
rm(regional_crime)
```

## Write Data
Now we can write our city-wide estimates:

```{r write-data}
write_csv(regions, path = here("data", "region", "region.csv"))
```
