---
title: "Neighborhood Crime Rate Plots, 2017 through 2019"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook creates a set of plots documenting changes in neighborhood level crime rates in St. Louis.

## Dependencies
This notebook requires the following `R` packages:

```{r load-packages, message=FALSE}
# tidyverse packages
library(dplyr)
library(ggplot2)
library(readr)

# other packages
library(here)
```

In addition, a custom plot theme is also loaded from `source/`:

```{r load-functions}
source(here("source", "cp_sequoiaTheme.R"))
```

## Load Data
We'll need the four data tables in `data/`:

```{r load-data, results='hide'}
city_wide <- read_csv(file = here("data", "city_wide.csv"))
homicide <- read_csv(file = here("data", "homicide.csv"))
part1 <- read_csv(file = here("data", "part1.csv"))
violent <- read_csv(file = here("data", "violent.csv"))
```

## Additional Data Cleaning
There are a few additional data cleaning steps to prepare our tables for plotting.

### Add County-Level Data
The county-level data are stored in a separate table (following tidy data principles), but we want them to appear as a separate line in our bar plots. We'll create individual tibbles to bind to the bottom of our neighborhood crime obects. Fist, crime:

```{r homicide-cleaning}
# create city-level tibble
city <- tibble(
  NHD_NUM = c(100),
  NHD_NAME = c("City of St. Louis"),
  total_pop = city_wide$total_pop[1],
  count_h_18 = city_wide$homicide_count[1],
  rate_h_18 = city_wide$homicide_rate[1],
  count_h_18p = city_wide$homicide_count[2],
  rate_h_18p = city_wide$homicide_rate[2],
  count_h_19p = city_wide$homicide_count[3],
  rate_h_19p = city_wide$homicide_rate[3],
  ab_change = city_wide$homicide_count[3]-city_wide$homicide_count[2]
)

# bind rows
homicide <- bind_rows(homicide, city)
```

Next, we'll clean the violent crime data in a similar fashion:

```{r violent-cleaning}
# create city-level tibble
city <- tibble(
  NHD_NUM = c(100),
  NHD_NAME = c("City of St. Louis"),
  total_pop = city_wide$total_pop[1],
  count_v_18 = city_wide$violent_count[1],
  rate_v_18 = city_wide$violent_rate[1],
  count_v_18p = city_wide$violent_count[2],
  rate_v_18p = city_wide$violent_rate[2],
  count_v_19p = city_wide$violent_count[3],
  rate_v_19p = city_wide$violent_rate[3],
  pct_change = (city_wide$violent_rate[3]-city_wide$violent_rate[2])/city_wide$violent_rate[2]*100
)

# bind rows
violent <- bind_rows(violent, city)
```

Finally, we'll clean the part 1 crime data:

```{r part-1-cleaning}
# create city-level tibble
city <- tibble(
  NHD_NUM = 100,
  NHD_NAME = "City of St. Louis",
  total_pop = city_wide$total_pop[1],
  count_p1_18 = city_wide$p1_count[1],
  rate_p1_18 = city_wide$p1_rate[1],
  count_p1_18p = city_wide$p1_count[2],
  rate_p1_18p = city_wide$p1_rate[2],
  count_p1_19p = city_wide$p1_count[3],
  rate_p1_19p = city_wide$p1_rate[3],
  pct_change = (city_wide$p1_rate[3]-city_wide$p1_rate[2])/city_wide$p1_rate[2]*100
)

# bind rows
part1 <- bind_rows(part1, city)
```

### Add Highlighting Indicators
We want to highlight the four "SLU" neighborhoods in addition to highlighting the City of St. Louis bar:

```{r highlight}
homicide %>%
  mutate(highlight = "a") %>%
  mutate(highlight = ifelse(NHD_NAME == "City of St. Louis", "c", highlight)) -> homicide

violent %>%
  mutate(highlight = "a") %>%
  mutate(highlight = ifelse(NHD_NAME == "City of St. Louis", "c", highlight)) -> violent

part1 %>%
  mutate(highlight = "a") %>%
  mutate(highlight = ifelse(NHD_NAME == "City of St. Louis", "c", highlight)) -> part1
```

### Remove City Tables
These are no longer needed:

```{r clean-up}
rm(city, city_wide)
```

## 2018 Full Year Plots
First, we'll create a series of plots just for the full 2018 data.

### Homicide

```{r homicide-2018-full}
ggplot(data = homicide, mapping = aes(x = reorder(NHD_NAME, rate_h_18), y = rate_h_18)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 6), breaks = seq(0, 6, .5)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Homicides, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (2018)",
    x = "",
    y = "Homicide Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nNeighborhoods with no bars have a homicide rate of '0'\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "homicide_18_full.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Violent Crime

```{r violent-2018-full}
ggplot(data = violent, mapping = aes(x = reorder(NHD_NAME, rate_v_18), y = rate_v_18)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 100), breaks = seq(0, 100, 10)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Violent Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (2018)",
    x = "",
    y = "Violent Crime Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nViolent crimes include homicides, rapes, aggrevated assaults, and robberies\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "violent_18_full.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Part 1 Crime

```{r part-1-2018-full}
ggplot(data = part1, mapping = aes(x = reorder(NHD_NAME, rate_p1_18), y = rate_p1_18)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 300), breaks = seq(0, 300, 50)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Part 1 Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (2018)",
    x = "",
    y = "Part 1 Crime Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPart 1 crimes include violent crimes (homicides, rapes, aggrevated assaults, and robberies) as well as crimes \n  against property (burlary, larceny, larceny of a vehicle, arson)\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "part1_18_full.png"), dpi = 500, height = 16, width = 12, units = "in")
```

## 2018 Partial Year Plots

### Homicide

```{r homicide-2018-partial}
ggplot(data = homicide, mapping = aes(x = reorder(NHD_NAME, rate_h_18p), y = rate_h_18p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 6), breaks = seq(0, 6, .5)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Homicides, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2018)",
    x = "",
    y = "Homicide Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "homicide_18_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Violent Crime

```{r violent-2018-partial}
ggplot(data = violent, mapping = aes(x = reorder(NHD_NAME, rate_v_18p), y = rate_v_18p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 80), breaks = seq(0, 80, 10)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Violent Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2018)",
    x = "",
    y = "Violent Crime Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nViolent crimes include homicides, rapes, aggrevated assaults, and robberies\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "violent_18_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Part 1 Crime

```{r part-1-2018-partial}
ggplot(data = part1, mapping = aes(x = reorder(NHD_NAME, rate_p1_18p), y = rate_p1_18p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 200), breaks = seq(0, 200, 25)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Part 1 Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2018)",
    x = "",
    y = "Part 1 Crime Rate",
  caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPart 1 crimes include violent crimes (homicides, rapes, aggrevated assaults, and robberies) as well as crimes \n  against property (burlary, larceny, larceny of a vehicle, arson)\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "part1_18_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

## 2019 Partial Year Plots

### Homicide

```{r homicide-2019-partial}
ggplot(data = homicide, mapping = aes(x = reorder(NHD_NAME, rate_h_19p), y = rate_h_19p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 6), breaks = seq(0, 6, .5)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Homicides, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2019)",
    x = "",
    y = "Homicide Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "homicide_19_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Violent Crime

```{r violent-2019-partial}
ggplot(data = violent, mapping = aes(x = reorder(NHD_NAME, rate_v_19p), y = rate_v_19p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 80), breaks = seq(0, 80, 10)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Violent Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2019)",
    x = "",
    y = "Violent Crime Rate",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nViolent crimes include homicides, rapes, aggrevated assaults, and robberies\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "violent_19_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

### Part 1 Crime

```{r part-1-2019-partial}
ggplot(data = part1, mapping = aes(x = reorder(NHD_NAME, rate_p1_19p), y = rate_p1_19p)) +
  geom_bar(stat = "identity", aes(fill = highlight)) + 
  scale_y_continuous(limits=c(0, 200), breaks = seq(0, 200, 25)) +
  coord_flip() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
  labs(
    title = "Part 1 Crimes, Rate per 1000 Residents",
    subtitle = "St. Louis, MO (January through August, 2019)",
    x = "",
    y = "Part 1 Crime Rate",
  caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPart 1 crimes include violent crimes (homicides, rapes, aggrevated assaults, and robberies) as well as crimes \n  against property (burlary, larceny, larceny of a vehicle, arson)\nPlot by Christopher Prener, Ph.D."
  ) +
  cp_sequoiaTheme(background = "white", base_size = 16) +
  theme(
    legend.position = ""
  )

ggsave(here("results", "part1_19_partial.png"), dpi = 500, height = 16, width = 12, units = "in")
```

## Change from 2018 to 2019

### Violent Crime

```{r violent-pct-change}
violent %>%
  filter(NHD_NUM != 40) %>%
  filter(NHD_NUM != 45) %>%
  ggplot(data = ., mapping = aes(x = reorder(NHD_NAME, pct_change), y = pct_change)) +
    geom_bar(stat = "identity", aes(fill = highlight)) + 
    scale_y_continuous(limits=c(-80, 120), breaks = seq(-80, 120, 20)) +
    coord_flip() +
    scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
    labs(
      title = "Percent Change in Violent Crime Rates per 1000 Residents",
      subtitle = "St. Louis, MO (January through August, 2018 to 2019)",
      x = "",
      y = "Percent Change",
      caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nViolent crimes include homicides, rapes, aggrevated assaults, and robberies\nWydown Skinker and Kings Oak removed due 0 violent crimes in 2018 or 2019\nPlot by Christopher Prener, Ph.D."
    ) +
    cp_sequoiaTheme(background = "white", base_size = 16) +
    theme(
      legend.position = ""
    )

ggsave(here("results", "violent_pctChange_partial.png"), dpi = 500, height = 16, width = 20, units = "in")
```

### Part 1 Crime

```{r part-1-pct-change}
part1 %>%
  ggplot(data = ., mapping = aes(x = reorder(NHD_NAME, pct_change), y = pct_change)) +
    geom_bar(stat = "identity", aes(fill = highlight)) + 
    scale_y_continuous(limits=c(-60, 60), breaks = seq(-60, 60, 20)) +
    coord_flip() +
    scale_fill_manual(values = c("#66C2A5", "#8DA0CB")) +
    labs(
      title = "Percent Change in Part 1 Crime Rates per 1000 Residents",
      subtitle = "St. Louis, MO (January through August, 2018 to 2019)",
      x = "",
      y = "Percent Change",
    caption = "Data via SLMPD and U.S. Census Bureau (2013-2017 5-year American Community Survey)\nPart 1 crimes include violent crimes (homicides, rapes, aggrevated assaults, and robberies) as well as crimes against property (burlary, larceny, larceny of a vehicle, arson)\nPlot by Christopher Prener, Ph.D."
    ) +
    cp_sequoiaTheme(background = "white", base_size = 16) +
    theme(
      legend.position = ""
    )

ggsave(here("results", "part1_pctChange_partial.png"), dpi = 500, height = 16, width = 20, units = "in")
```
